---
title: "Five Minute Baby Steps"
author: "Adam O'Dell"
date: "Tuesday, July 14, 2015"
output: html_document
---

First things, first: let's load the data into a dataframe (df).

```{r}

df <- read.csv("activity.csv")

```

Next, we'll check to make sure the dimensions are what we expect (17568 x 3):

```{r}
dim(df)
```

And double-check the column/variable heading:

```{r}
names(df)
```

And glance at the "head" of the dataset:

```{r}
head(df)
```

All looks well. Now, let's do some analysis.

First, we want to calculate the Total (i.e. sum) number of steps taken each day (TPD = TotalperDay).

```{r}

TPD <- aggregate(. ~ date, data =df, sum)
TPD
```

Visual learners, like myself, will prefer to see that data as a histogram:

```{r}
Steps_Per_Day <- TPD$steps
hist(Steps_Per_Day)
```

Next, let's compute two simple summary statistics: mean and median.

```{r}

steps_mean <- mean(Steps_Per_Day)
steps_median <- median(Steps_Per_Day)

steps_mean
steps_median
```

Now, let's "drill down" and look more closely at each 5-minute interval.

Just as we look at statistics for each data, let's do the same for each interval (API = AveragePerInterval).

```{r}

API <- aggregate(.~interval, data=df, mean)

API
```

Again, a VISUAL will help...

```{r}

plot(API$interval, API$steps, type = "l")
```

Visually, it's easy to see the "spike" in steps.

But if someone asked you, "Which specific interval averages the most steps?"

We'd have to calculate that...

```{r}

MaxInterval <- API[which(API$steps == max(API$steps)),]
MaxInterval

```

OK, so Interval #835 is the winner, with 206.2 steps on average.

Now, let's move on to something altogether different.

There seem to be some missing values (i.e. "NA"s) in our dataset. See...

```{r}
head(df)
```

Let's address those missing values.


First, let's figure out how many rows within our dataset contain at least one missing value (i.e. "NA").

```{r}
TotalNAs <- sum(is.na(df))
TotalNAs
```

Wow, 2304 missing values... that's a lot.

Let's now devise a simple strategy for replacing those missing values. There are many ways to do this.

I'll choose this one: replace any NA value with the value equal to the average value of the interval in question.  

```{r}

df_new <- df

df_new$steps <- ifelse(is.na(df_new$steps) == TRUE, API$steps[API$interval %in% df_new$interval], df_new$steps)

```

Let's see if that worked...

```{r}
head(df_new)
tail(df_new)
sum(is.na(df_new))
```

Looks good to me!

Now, let's make a histogram of the new dataframe, which is equivalent to the original dataframe, only with all the NAs replaced.

```{r}
TotalPerDay <- aggregate(. ~ date, data=df_new, sum)

Steps_Per_Day <- TotalPerDay$steps
hist(Steps_Per_Day)
```

Visually, that histogram looks just like the original histogram of Total Steps Per Day, which had missing NAs value.

But let's be more precise. Calculating the mean and median of our new dataset, with NAs replaced, will help.

```{r}
new_mean <- mean(Steps_Per_Day)
new_median <- median(Steps_Per_Day)

new_mean
new_median
```
Compared to the "original" values:

```{r}
steps_mean
steps_median
```

The mean has not changed at all. And the median has changed only ever so slightly (basically, it hasn't changed).

Sadly, it seems our replacement of the NAs had no effect... it was a waste of time (except I'm learning R!)

Now, let's explore one last thing.

I wonder if people take more steps on Weekdays versus Weekend Days.

First, we have to create a new variable - one that defines the date variable as either: "Weekday" or "Weekend". That sounds like a "factor" variable with two levels.

We'll start this work on our "NA-replaced" dataset (i.e. df_new)

```{r}
df_new$date <- as.Date(df_new$date)

weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')

df_new$wDay <- factor((weekdays(df_new$date) %in% weekdays1), levels = c(FALSE,TRUE), labels=c('Weekend','Weekday'))

```

Let's make sure our new variable is a "factor" variable:

```{r}
class(df_new$wDay)
```

And glance at the head of our new dataset:

```{r}
head(df_new)
```

Good, all looks well!

Next, let's subset this dataset based on our Weekday/Weekend factor variable:

```{r}

weekend_set <- subset(df_new, wDay == "Weekend")

weekday_set <- subset(df_new, wDay == "Weekday")

```

Now we can compute the Average Steps per Interval, for each subset:

```{r}

API_weekend <- aggregate(. ~ interval, data = weekend_set, mean)
API_weekend$wDay <- factor("Weekend")

API_weekday <- aggregate(. ~ interval, data = weekday_set, mean)
API_weekday$wDay <- factor("Weekday")
```

Let's recombine those subsets, to make a Panel Plot easier.

```{r}
new_API <- rbind(API_weekend,API_weekday)
```

And, finally, make a Panel plot... comparing the diference in Average Steps per Interval for Weekdays versus Weekends:

```{r}
library(lattice)
xyplot(steps ~ interval | wDay,
        data = new_API,
        type = "l",
        xlab = "Interval",
        ylab = "Average Steps",
        layout = c(1,2))
```

Interesting!

It seems that on Weekdays, more steps are taken between the 500 and 1000 intervals.

But on Weekends, more steps are taken between the 1000 and 2000 intervals.

Although, overall, the patterns of activity are very similar on both Weekdays and Weekends.

This was a fun investigation of data!

See you next time...
